---
title: "Comparing R packages"
subtitle: "`packagemetrics` figure"
author: "Caroline Wendt"
date: "8/21/2020"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      message = FALSE, error = FALSE, warning = FALSE)
```

## Installation
```{r}
#devtools::install_github("ropenscilabs/packagemetrics")
```

## Packages
```{r}
library(packagemetrics)
```

## Comparing packages for data visualization and graphics
*Note:* One main use for packagemetrics is to compare similar/overlapping packages. So we demonstrate an example.

```{r}
# get info on packages
pkgs <- package_list_metrics(c("ggplot2",
                                    "ggforce",
                                    "RGraphics",
                              "misc3d",
                              "scatterplot3d",
                              "rgl",
                              "onion",
                               "plotly", 
                               "flexdashboard",
                               "shiny",
                               "dygraphs", 
                               "leaflet", 
                               "patchwork",
                               "colourpicker",
                               "RColorBrewer",
                               "paletteer",
                               "colorspace",
                               "vcd",
                               "dichromat",
                               "highcharter",
                               "echarts4r",
                               "geofacet",
                               "quantmod",
                               "tidyquant",
                               "patchwork"))
dplyr::glimpse(pkgs)
```

## Create a nice `metrics_table`

We will use this figure as an example in the Ten Rules paper.

```{r}
# data frame
pkg_df <- package_list_metrics(list("ggplot2",
                                    "ggforce",
                                    "RGraphics",
                              "misc3d",
                              "scatterplot3d",
                              "rgl",
                              "onion",
                               "plotly", 
                               "flexdashboard",
                               "shiny",
                               "dygraphs", 
                               "leaflet", 
                               "patchwork",
                               "colourpicker",
                               "RColorBrewer",
                               "paletteer",
                               "colorspace",
                               "vcd",
                               "dichromat",
                               "highcharter",
                               "echarts4r",
                               "geofacet",
                               "quantmod",
                               "tidyquant",
                               "patchwork")) 
table <- metrics_table(pkg_df)
```

```{r}
# html table widget
table_html <- formattable::as.htmlwidget(table, width = 1200, height = NULL)

path <- htmltools::html_print(table_html, background = "white", viewer = NULL)

url <- paste0("file:///", gsub("\\\\", "/", normalizePath(path)))

webshot::webshot(url,
                 file = "inst/image/table.png",
                 selector = ".formattable_widget",
                 delay = 0.2)
```

## Refreshing data sources

If the cached CRAN data gets out of date, call `get_cran(forget = TRUE)` to force refreshing.  This will fetch the latest results from CRAN and cache them for use going forward.

GitHub data is scraped at the time of the call to `package_list_metrics`.
